Data Migration under a Changing Schema in Ampersand

### 2023-10-18 2022Migration/articleMigrationFACS.tex
Introduce the Proof Obligations

   \begin{definition}[Schema]
   A schema is a tuple $\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ that satisfies:
\begin{align}
   \declare{n}{A}{B}\in\rels&~\Rightarrow~ A\in\concepts\,\wedge\, B\in\concepts
   \label{eqn:relationsIntroduceConcepts}\\
   u\in\rules\cup\transactions\cup\busConstraints\ \wedge\ \sign{u}=\pair{A}{B}&~\Rightarrow~A\in\concepts\,\wedge\, B\in\concepts\label{eqn:invariant-has-type}\\
   (\declare{n}{A}{B}\mapsfrom t)\in\mathcal\transactions&~\Rightarrow~ \declare{n}{A}{B}\in \rels\label{eqn:enforcement-has-type}
\end{align}
   \end{definition}
   Requirements~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:invariant-has-type} ensure that concepts mentioned in relations and in the signature of rules are part of the schema.
   Requirement~\ref{eqn:enforcement-has-type} ensures the enforced relation of a transaction is declared in the schema. 
   When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, $\transactions_{\schema}$, and $\busConstraints_{\schema}$
   for $\concepts$, $\rels$, $\rules$, $\transactions$, and $\busConstraints$ corresponding to $\schema = \quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$.

\subsection{Information Systems}
\label{sct:Information Systems}
   Let us now define information systems by enumerating the requirements.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\pair{\dataset}{\schema}$, in which
\begin{itemize}
\item $\dataset=\pair{\triples}{\instance}$ is a data set (so it must satisfy equation~\ref{eqn:wellTypedEdge}). We write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$;
\item Triples in the data set must have their relation mentioned in the schema:
   \begin{eqnarray}
      \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}
   \end{eqnarray}
\item $\schema=\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}).
\item All violations must have a type, which follows from  (\ref{eqn:wellTypedViolation}).
\item All blocking invariants and transactions must remain satisfied:
   \begin{align}
      \forall u\in\rules\cup\transactions&.~\viol{u}{\dataset}=\emptyset
   \label{eqn:satisfaction}
   \end{align}
\item Transactions remain satisfied by adding violations to a specific relation (\ref{eqn:transaction}).
\end{itemize}
\end{definition}
   % A \define{role} is a name that identifies a group of users.
   % It serves as a placeholder for a person or a machine (i.e. an actor) that works with the data set (i.e. create, read, update, or delete triples).
   % The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.

@@

   Then we define the migration system $\migrsys$ as follows:
   Let $\pair{\dataset}{\schema}$ be the existing system.
   Let $\pair{\dataset'}{\schema'}$ be the desired system in its initial state.
\begin{enumerate}
\item We take a disjoint union of the data sets by relabeling relation names:
   \begin{align}
      \dataset_\migrsys={}&\overleftarrow{\dataset}\cup\overrightarrow{\dataset'}
   \end{align}
\item We create transactions to copy equal relations:
For every relation $r$ shared by the existing and desired systems, we generate a helper relation: ${\tt copy}_r$, and a transaction to produce its population.
We also generate a transaction to populate the relation $\overrightarrow{r}$ (in the desired system).
% \begin{verbatim}
%    RELATION copied_r[A*B]
%    RELATION new_r[A*B] [UNI]
%    ENFORCE copied_r >: new_r /\ old_r
%    ENFORCE new_r >: old_r - copied_r
% \end{verbatim}
   \begin{align}
      \rels_1={}&\{{\tt copy}_r\mid r \in \rels_{\schema'}\cap\rels_{\schema}\}\\
      \transactions_1={}&\{{\tt copy}_r\mapsfrom \popF{\overrightarrow{r}\cap\overleftarrow{r}} \mid r \in\rels_1\}\cup\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r}-{\tt copy}_r} \mid r \in\rels_1\}
   \end{align}
   The helper relation is needed to ensure that the copying process terminates.
\item The new blocking invariants are $\rules_{\schema'}-\rules_{\schema}$.
   For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, to register all violations that are fixed.
   We also need a transaction to produce its population:
% \begin{verbatim}
%    RELATION fixed_TOTr[A*A]
%    ENFORCE fixed_TOTr >: I /\ new_r;new_r~
% \end{verbatim}
   \begin{align}
      \rels_2={}&\{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
      \transactions_2={}&\{{\tt fixed}_u\mapsfrom\lambda\dataset.~\viol{\overleftarrow{u}}{\dataset}-\viol{\overrightarrow{u}}{\dataset}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
   \end{align}
\item For each new blocking invariant $u$, we generate another blocking invariant $v$ in the migration system that blocks fixed violations from recurring:
% \begin{verbatim}
%    RULE Block_TOTr : fixed_TOTr |- new_r;new_r~
%    MESSAGE "Relation r[A*B] must remain total."
%    VIOLATION (TXT "Atom ", SRC I, TXT " must remain paired with an atom from B.")
% \end{verbatim}
   \begin{align}
      \rules_\text{block}={}&\{v\ 
      \begin{array}[t]{l}
         \text{\bf where}\\
         \sign{v}=\sign{u}\\
         \viol{v}{\dataset}=\viol{\overrightarrow{u}}{\dataset}\cap{\tt fixed}_u
      \end{array}\\
      &\mid u\in\rules_{\schema'}-\rules_{\schema}\}\notag
   \end{align}
   Note that the blocking invariants from $\schema'\cap\schema$ are also used in $\migrsys$ to ensure continuity.
\item To signal users that there are violations that need to be fixed, we generate a business constraint for each new blocking invariant $u$:
% \begin{verbatim}
%    ROLE User MAINTAINS TOTr
%    RULE TOTr : I - fixed_TOTr |- new_r;new_r~
%    MESSAGE "Please, make relation r[A*B] total."
%    VIOLATION (TXT "Fix ", SRC I, TXT " by pairing it with an (any) atom from B.")
% \end{verbatim}
   \begin{align}
      \busConstraints_\text{fix}={}&\{v\ 
      \begin{array}[t]{l}
         \text{\bf where}\\
         \sign{v}=\sign{u}\\
         \viol{v}{\dataset}=\viol{\overrightarrow{u}}{\dataset}-{\tt fixed}_u
      \end{array}\\
      &\mid u\in\rules_{\schema'}-\rules_{\schema}\}\notag
   \end{align}
\item Let us combine the above into a single migration schema:
   \begin{align}
      \schema_\migrsys=\langle{}&\concepts_\dataset\cup\concepts_{\dataset'},\\
      &\overleftarrow{\rels_{\schema}}\cup\overrightarrow{\rels_{\schema'}}\cup\rels_1\cup\rels_2,\notag\\
      &\rules_\text{block}\cup(\rules_{\schema}\cap\rules_{\schema'}),\notag\\
      &\transactions_1\cup\transactions_2,\notag\\
      &\busConstraints_\text{fix}\cup\busConstraints_{\schema'}\rangle\notag
   \end{align}
\end{enumerate}
   The migration engineer can switch all traffic to the desired system
   after resolving the violations that prohibit deploying the desired system.
   That is the case when violations of new invariants on the old population have all been fixed:
% \begin{verbatim}
%    ROLE User MAINTAINS CleanUp_TOTr
%    RULE CleanUp_TOTr : V[ONE*A] ; (I - fixed_TOTr) ; V[A*ONE]
%    MESSAGE "Now you can remove the migration system because relation r[A*B] is total."
% \end{verbatim}
\begin{equation}
   \forall u\in\rules_{\schema'}-\rules_{\schema}.\ \viol{\overleftarrow{u}}{\dataset_\migrsys}={\tt fixed}_u
\end{equation}
   After this, the migration engineer can remove the migration system and the old system.

\subsection{Validation}
   We first observe that $\migrsys = \pair{\dataset_\migrsys}{\schema_\migrsys}$ is a system after application of the transactions:
   the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and transactions from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.
   Hence, the MoT can happen after the transactions have been applied, giving us $\migrsys'$.
   The event $\migrsys \xrightarrow[\rels_2 \cup \rels_3]{\overrightarrow\rels_{\schema'}}\migrsys'$ represents the MoT.

>>>>

   \begin{definition}[Schema]
   A schema is a tuple $\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ that satisfies:
\begin{align}
   \declare{n}{A}{B}\in\rels&~\Rightarrow~ A\in\concepts\,\wedge\, B\in\concepts
   \label{eqn:relationsIntroduceConcepts}\\
   u\in\rules\cup\transactions\cup\busConstraints\ \wedge\ \sign{u}=\pair{A}{B}&~\Rightarrow~A\in\concepts\,\wedge\, B\in\concepts\label{eqn:invariant-has-type}\\
   (\declare{n}{A}{B}\mapsfrom t)\in\mathcal\transactions&~\Rightarrow~ \declare{n}{A}{B}\in \rels\label{eqn:enforcement-has-type}
\end{align}
   \end{definition}
   Requirements~\ref{eqn:relationsIntroduceConcepts} and~\ref{eqn:invariant-has-type} ensure that concepts mentioned in relations and in the signature of rules are part of the schema.
   Requirement~\ref{eqn:enforcement-has-type} ensures the enforced relation of a transaction is declared in the schema. 
   When clarity is needed, we write $\concepts_{\schema}$, $\rels_{\schema}$, $\rules_{\schema}$, $\transactions_{\schema}$, $\busConstraints_{\schema}$
   for $\concepts$, $\rels$, $\rules$, $\transactions$, $\busConstraints$ corresponding to $\schema = \quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$.

\subsection{Information Systems}
\label{sct:Information Systems}
   Let us now define information systems by enumerating the requirements.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\pair{\dataset}{\schema}$, in which
\begin{itemize}
\item $\dataset=\pair{\triples}{\instance}$ is a data set (so it must satisfy equation~\ref{eqn:wellTypedEdge}).
   We write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$ if needed;
\item Triples in the data set must have their relation mentioned in the schema:
   \begin{eqnarray}
      \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}
   \end{eqnarray}
\item $\schema=\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}).
\item All violations must have a type, which follows from  (\ref{eqn:wellTypedViolation}).
\item All blocking invariants and transactions must remain satisfied:
   \begin{align}
      \forall u\in\rules\cup\transactions&.~\viol{u}{\dataset}=\emptyset
   \label{eqn:satisfaction}
   \end{align}
\item Transactions remain satisfied by adding violations to a specific relation (\ref{eqn:transaction}).
\end{itemize}
\end{definition}
   % A \define{role} is a name that identifies a group of users.
   % It serves as a placeholder for a person or a machine (i.e. an actor) that works with the data set (i.e. create, read, update, or delete triples).
   % The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.

@@

   Then we define the migration system $\migrsys$ as follows:
   Let $\pair{\dataset}{\schema}$ be the existing system.
   Let $\pair{\dataset'}{\schema'}$ be the desired system in its initial state.
\begin{enumerate}
\item We take a disjoint union of the data sets by relabeling relation names:
   \begin{align}
      \dataset_\migrsys={}&\overleftarrow{\dataset}\cup\overrightarrow{\dataset'}
   \end{align}
\item We create transactions to copy equal relations:
For every relation $r$ shared by the existing and desired systems, we generate a helper relation: ${\tt copy}_r$, and a transaction to produce its population.
We also generate a transaction to populate the relation $\overrightarrow{r}$ (in the desired system).
% \begin{verbatim}
%    RELATION copied_r[A*B]
%    RELATION new_r[A*B] [UNI]
%    ENFORCE copied_r >: new_r /\ old_r
%    ENFORCE new_r >: old_r - copied_r
% \end{verbatim}
   \begin{align}
      \rels_1={}&\{{\tt copy}_r\mid r \in \rels_{\schema'}\cap\rels_{\schema}\}\\
      \transactions_1={}&\{{\tt copy}_r\mapsfrom \popF{\overrightarrow{r}\cap\overleftarrow{r}} \mid r \in\rels_1\}\cup\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r}-{\tt copy}_r} \mid r \in\rels_1\}
   \end{align}
   The helper relation is needed to ensure that the copying process terminates,
   which is the case when:
   \begin{align}
      \forall r\in\rels_1.~\overleftarrow{r}={\tt copy}_r\label{eqn:copyingTerminates}
   \end{align}

\item The new blocking invariants are $\rules_{\schema'}-\rules_{\schema}$.
   For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, to register all violations that are fixed.
   We also need a transaction to produce its population:
% \begin{verbatim}
%    RELATION fixed_TOTr[A*A]
%    ENFORCE fixed_TOTr >: I /\ new_r;new_r~
% \end{verbatim}
   \begin{align}
      \rels_2={}&\{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
      \transactions_2={}&\{{\tt fixed}_u\mapsfrom\lambda\dataset.~\viol{\overleftarrow{u}}{\dataset}-\viol{\overrightarrow{u}}{\dataset}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
   \end{align}
   The helper relation is needed to ensure that the fixing of violations terminates,
   which is the case when:
   \begin{align}
      \forall u\in\rules_{\schema'}-\rules_{\schema}.~\viol{\overrightarrow{u}}{\dataset}={\tt fixed}_u\label{eqn:fixingTerminates}
   \end{align}
\item For each new blocking invariant $u$, we generate another blocking invariant $v$ in the migration system that blocks fixed violations from recurring:
% \begin{verbatim}
%    RULE Block_TOTr : fixed_TOTr |- new_r;new_r~
%    MESSAGE "Relation r[A*B] must remain total."
%    VIOLATION (TXT "Atom ", SRC I, TXT " must remain paired with an atom from B.")
% \end{verbatim}
   \begin{align}
      \rules_\text{block}={}&\{v\ 
      \begin{array}[t]{l}
         \text{\bf where}\\
         \sign{v}=\sign{u}\\
         \viol{v}{\dataset}=\viol{\overrightarrow{u}}{\dataset}\cap{\tt fixed}_u
      \end{array}\\
      &\mid u\in\rules_{\schema'}-\rules_{\schema}\}\notag
   \end{align}
   Note that the blocking invariants from $\schema'\cap\schema$ are also used in $\migrsys$ to ensure continuity.
\item To signal users that there are violations that need to be fixed, we generate a business constraint for each new blocking invariant $u$:
% \begin{verbatim}
%    ROLE User MAINTAINS TOTr
%    RULE TOTr : I - fixed_TOTr |- new_r;new_r~
%    MESSAGE "Please, make relation r[A*B] total."
%    VIOLATION (TXT "Fix ", SRC I, TXT " by pairing it with an (any) atom from B.")
% \end{verbatim}
   \begin{align}
      \busConstraints_\text{fix}={}&\{v\ 
      \begin{array}[t]{l}
         \text{\bf where}\\
         \sign{v}=\sign{u}\\
         \viol{v}{\dataset}=\viol{\overrightarrow{u}}{\dataset}-{\tt fixed}_u
      \end{array}\\
      &\mid u\in\rules_{\schema'}-\rules_{\schema}\}\notag
   \end{align}
\item Let us combine the above into a single migration schema:
   \begin{align}
      \schema_\migrsys=\langle{}&\concepts_\dataset\cup\concepts_{\dataset'},\\
      &\overleftarrow{\rels_{\schema}}\cup\overrightarrow{\rels_{\schema'}}\cup\rels_1\cup\rels_2,\notag\\
      &\rules_\text{block}\cup(\rules_{\schema}\cap\rules_{\schema'}),\notag\\
      &\transactions_1\cup\transactions_2\cup\transactions_{\schema'},\notag\\
      &\busConstraints_\text{fix}\cup\busConstraints_{\schema'}\rangle\notag
   \end{align}
\end{enumerate}
   The migration engineer can switch all traffic to the desired system
   after resolving the violations that prohibit deploying the desired system.
   That is the case when violations of new invariants on the old population have all been fixed:
% \begin{verbatim}
%    ROLE User MAINTAINS CleanUp_TOTr
%    RULE CleanUp_TOTr : V[ONE*A] ; (I - fixed_TOTr) ; V[A*ONE]
%    MESSAGE "Now you can remove the migration system because relation r[A*B] is total."
% \end{verbatim}
\begin{equation}
   \forall u\in\rules_{\schema'}-\rules_{\schema}.\ \viol{\overleftarrow{u}}{\dataset_\migrsys}={\tt fixed}_u
\end{equation}
   After this, the migration engineer can remove the migration system and the old system.

\subsection{Proof Obligations}
\begin{enumerate}
\item Assume that $\infsys$ and $\infsys'$ are information systems, so they satisfy definition~\ref{def:information system}.
      This means that equations~\ref{eqn:wellTypedEdge}, \ref{eqn:wellTypedViolation}, \ref{eqn:transaction}, \ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}, \ref{eqn:define R}, \ref{eqn:satisfaction} hold for both systems.
\item Prove that $\migrsys=\pair{\dataset_\migrsys}{\schema_\migrsys}$ is an information system,
      so the migration engineer can deploy the migration system.
\item Prove that $\overrightarrow{r}$ contains all triples from the existing data set in a finite time after deploying $\migrsys$,
      to ensure that the existing data is migrated to the desired system.
\item Prove that equation~\ref{eqn:satisfaction} holds for $\migrsys$ when the copying of relations is done (just before MoT),
      to ensure that a valid information system is presented to users at the moment of transition.
\item Prove that equation~\ref{eqn:copyingTerminates} represents the condition for terminating the copying process.
\item Prove that equation~\ref{eqn:fixingTerminates} represents the condition for terminating the fixing of violations.
\item Prove that the migration engineer can switch the ingress from the existing system to the migration system
      when equations~\ref{eqn:copyingTerminates} and~\ref{eqn:fixingTerminates} are true,
      so the computer can let the migration engineer know that it is safe to switch.
\item Assuming that the business can fix each violation in finite time,
      prove that the number of violations to be fixed by the business is finite,
      and that it decreases monotonically, so the business can reach the MoC in finite time.
\item Prove that $\rules_\text{block}$ is free of violations, so it is redundant after the MoC.
\item Prove that $\transactions_1$ will never more add any violations to any ${\tt copy}_u$ at the MoC, so $\rels_1$ and $\transactions_1$ are redundant after the MoC.
\item Prove that $\transactions_2$ will never more add any violations to any ${\tt fixed}_u$ at the MoC, so that $\rels_2$ and $\transactions_2$ are redundant after the MoC.
\item Prove that $\busConstraints_\text{fix}$ is free of violations, so it is redundant after the MoC.
\item Prove that $\concepts_{\schema}$, $\rels_{\schema}$, and $\rules_{\schema}$ are redundant after the MoC.
\item Prove that, given the redundancies proven above, the migration system is equivalent to $\infsys'$ after the MoC,
      so the migration engineer can switch the ingress to the desired system and remove the existing system and the migration system.
\end{enumerate}
\subsection{Validation}
   We first observe that $\migrsys = \pair{\dataset_\migrsys}{\schema_\migrsys}$ is a system after application of the transactions:
   the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and transactions from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.
   Hence, the MoT can happen after the transactions have been applied, giving us $\migrsys'$.
   The event $\migrsys \xrightarrow[\rels_2 \cup \rels_3]{\overrightarrow\rels_{\schema'}}\migrsys'$ represents the MoT.
